---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First, I want to load the package [dplyr][1] for data wrangling and the 
package [ggplot2][2] for making plots.  I will also set the option to remove scientific 
notation from the output and round decimals to 4 spaces.

```{r}
library(dplyr)
library(ggplot2)
options(scipen=999, digits = 4)
```

Next, I'll load the activity.csv to a data.frame called "data."
```{r}
filename <- "activity.csv"
data <- read.csv(filename, stringsAsFactors = FALSE)
```

## What is mean total number of steps taken per day?

We want to find the total number of steps taken per day in order to calculate 
the mean, so I used *aggregate()* to sum the **steps** column for each unique **date.**

This is what the data looks like:

```{r}
stepsDay <- aggregate(steps ~ date, data = data, sum, na.action = na.omit)
qplot(steps, data=stepsDay, geom="histogram", binwidth = 500)
```

Now I can calculate the average number of steps per day:
```{r}
mean <- mean(stepsDay$steps)
median <- median(stepsDay$steps)
```

**Mean** steps per day: `r mean`

**Median** steps per day: `r median`


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?

### Links
[1]: http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html
[2]: http://cran.r-project.org/web/packages/ggplot2/index.html